scTOP: physics-inspired order parameters for cellular identification and visualization

### 2023-07-21 manuscript/scTOP response to reviewers/review_response.tex

          \begin{tabular}{ |p{2.5in}|p{2.5in}|} 
             \hline
             scTOP space & PCA space\\ 
             \hline
             \hline
             Defined by an input reference basis of known cell types & Calculated by PCA by finding the orthogonal vectors (principal components) that best explain the variation in data  \\ 
             \hline
             Can be non-orthogonal (e.g. can have highly-correlated cell types) & Orthogonal by definition  \\ 
             \hline
             Dimensions have clear biological interpretation: alignment with cell types & Dimensions need to be closely analyzed to be interpreted \\
             \hline
         \end{tabular}
	\end{changes}
\end{revresponse}

\begin{revcomment}
    The authors say scTOP does not require data harmonization and 
is able to distinguish between technical and biological noise, but 
couldn’t batch effects between the reference dataset and the sample 
dataset effect the results? Can you use some dataset to illustrate this,
 there are few discussed in the several batch correction papers?
\end{revcomment}
\begin{revresponse}
    scTOP is able to identify cell types and give similar score distributions even across data sets. To demonstrate this, we analyzed dataset 4 from Tran et al. \cite{tran2020benchmark}. This data set contains pancreatic data from 5 different sources. We have added the following section to the Supplementary Information:

    \begin{changes}      
        \subsection*{Batch effects}
        scRNA-seq data produced using different sequencing platforms from different experiments may introduce variation in the form of batch effects. Because the pre-processing step in scTOP prioritizes the relative rather than absolute expression of genes, batch effects are mitigated.

        Tran et al. \cite{tran2020benchmark} consolidated datasets from different experiments for the purposes of benchmarking various batch-effect correction methods. One of these is a dataset containing scRNA-seq human pancreatic data from five different sources \cite{baron2016single} \cite{muraro2016single} \cite{segerstolpe2016single} \cite{wang2016single} \cite{xin2016rna}, using various sequencing technologies. We used scTOP to analyze these datasets. Figure \ref{batch effects} illustrates that scTOP is not significantly affected by differences between batches. Because the Baron et al. dataset had the most cells per cell type, we used it to create a reference basis consisting of alpha, beta, ductal, acinar, delta, gamma, stellate, and endothelial gene expression profiles. We projected cells from the other four experiments onto the Baron reference basis. The top row shows kernel density estimate (KDE) plots for cells that match the reference cell type (e.g. alpha scores for alpha cells) with solid lines, and cells that do not match the reference type but have higher scores than the other non-matching types (e.g. alpha scores for beta cells) with dashed lines. The bottom row shows box-and-whisker plots for the cells that match the reference type.

        Despite coming from different sources, the distributions of scores are similar. There is also a clear bimodal distribution for cells that match the reference type and cells that don't. This bimodality is a clear indication that batch effects don't cause misidentification.

        According to Ziegenhain et al. \cite{ziegenhain2017comparative}, Smart-seq2 detects the most genes per cell when compared to other sequencing methods of that time. This decrease in dropouts may explain why the Segerstolpe data (which is the only experiment that used Smart-seq2) has slightly higher projections than the other datasets.
    \end{changes}

    \begin{figure}
	\centering
    \includegraphics[scale=0.95]{figs/batch effects.pdf}
	\caption{Pancreatic data from different experiments using different scRNA-seq technologies. The reference basis uses data from Baron et al. \cite{baron2016single}, and the cells that are projected onto this basis are from Muraro et al. \cite{muraro2016single}, Segerstolpe et al. \cite{segerstolpe2016single}, Wang et al. \cite{wang2016single}, and Xin et al. \cite{xin2016rna} (A) Kernel density estimate (KDE) plot of alpha scTOP scores for alpha cells (solid) and beta cells (dashed), and a box-and-whisker plot of alpha scTOP scores for alpha cells. (B) KDE plot of beta scTOP scores for beta cells (solid) and delta cells (dashed), and a box-and-whisker plot of beta scTOP scores for beta cells. }
	\label{batch effects}
    \end{figure}
\end{revresponse}

\begin{revcomment}
    Authors mention that pre-processing to generate z-scores is essential.
 It is not clear why that should be, at least authors don’t explain it. 
For example, does a simpler, more intuitive normalization per cell of 
converting counts to fractions, work? I think there are many schemes of 
normalizations that have been proposed and having a clear analysis and 
discussion around that is essential for “users” who may be interested in
 using scTOP.
\end{revcomment}
\begin{revresponse}
    To address this comment, we have added the following subsection under the ``Preprocessing'' section of the Supplementary Information.

    \begin{changes}
    \subsection*{Preprocessing rationale}

    In order to mitigate batch effects, the preprocessing step in scTOP takes each cell and ranks each gene according to expression. These ranks are converted into percentiles, which are then transformed into z-scores assuming a normal distribution.

    By ranking the genes before converting to z-scores, we are ignoring the absolute values of RNA counts and instead focusing on the relative values. Some sequencing platforms may measure many more counts than others; although we can't recover the unmeasured counts that are lost via dropouts, we can remove the effect of differences in counts by ranking. We assume that each cell contains similar numbers of mRNA and that the actual number of RNA doesn't matter as much as the relative expression of different genes. For example, the highest-expressed gene in one cell should be comparable to the highest-expressed gene in another cell, regardless of RNA count. Ignoring the absolute values of RNA counts reduces the differences caused by different scRNA-seq technologies and experiments, as shown by the ``Batch effects'' section.

    The second preprocessing step is to convert the ranks to z-scores, assuming a normal distribution. This step results in a mean of zero and standard deviation of one. In data analysis, it is beneficial to standardize the data before comparing across datasets because this prevents a disproportionate bias towards features with large values. In the case of scTOP, having a mean of zero and a standard deviation of one is especially important because the underlying math relies on the dot product of two uncorrelated vectors to be zero and the magnitude of a vector to be one.

    By ranking and z-scoring, we remove undue weighting of genes according to magnitude of expression, instead prioritizing the relative expression.
    \end{changes}

\end{revresponse}

\begin{revcomment}
    About lineages, this is perhaps the most important application of 
scTOP as it brings out something beyond cell types that was the input to
 the system. The authors have some discussion of how having fewer cell 
types make it harder to map out the space and hence the lineages. But 
many applications where one is interested in cell fate dynamics of only a
 handful of cell types, sometimes one cell type going to two, is central
 to understand the underlying genetic logic of that drives the cascade 
of “bifurcations”. This feels like a big limitation, and I am curious if
 authors to say a little more on this, perhaps again by taking a 
concrete example, either with synthetic or real data.
\end{revcomment}
\begin{revresponse}
    One limitation of scTOP is that a certain number of cells are needed per cell type in order to create an accurate expression profile for that cell type. While having more cell types can indeed help map out the cell type subspace, most cell types can be removed from the reference basis and scTOP scores are still able to distinguish the remaining types (see Supplementary Information section ''Robustness of scTOP``, which shows the result of having few cell types in the reference basis).

    scTOP is able to visualize the cell fate dynamics of one cell type going to two. A concrete example can be found in the hematopoiesis lineage tracing data. In this dataset, clonal families with three time points reveal the dynamics of hematopoietic stem cells differentiating into mature blood cells. Figure 6 shows the scTOP scores of individual clonal families (i.e. sister cells that originated from the same stem cell) over time. One of these plots shows the case of one cell type going to two: progenitor cells going to neutrophil and monocytes. The plot with basophils, neutrophils, and eosinophils shows one cell type going to three other cell types. Another example of one cell type going to two is shown by the AT1/AT2 plots, where a lung alveolar progenitor cell differentiates into either AT1 or AT2.

    The biggest limitation of scTOP is the case where the cell type of a sample of interest is not included in the reference basis. In this case, all the corresponding scTOP scores will be low, unless the cell type of the sample is highly correlated with a cell type in the basis. For this reason, the scTOP scores may be sensitive to change when highly-correlated cell types are added or removed.
\end{revresponse}

\begin{revcomment}
    Also, on lineages, could some analysis be provided to see if expected gene expression changes happen along the directions of cell fate changes? This would provide some confidence on the predicted cell fate 
dynamics.
\end{revcomment}
\begin{revresponse}
    Figure 5 of the main text shows the development of mouse alveolar cells. The scTOP scores for AT1 and AT2 are shown for each time point of this time series data, and each cell is colored by the expression of AT1 or AT2 marker genes. These marker genes are the genes that are expected to change and become highly expressed when AT1 or AT2 fates are adopted. As shown by the plots, expression of these marker genes occurs along the directions of cell fate indicated by scTOP. The expected gene expression changes do occur with the dynamics visualized by scTOP.
\end{revresponse}

\begin{revcomment}
    You should explicitly mention “xi” is the basis in methods 
section II A. I don’t think its explicitly mentioned anywhere and would 
be useful for readers not familiar with these types of methods.
\end{revcomment}
\begin{revresponse}
    To address this comment, we have added the following line:

    \begin{changes}
         We refer to $\xi$ as the reference basis.
    \end{changes}
\end{revresponse}

\begin{revcomment}
    Figure 2 a,b,c are referred to in the text but the figure does not contain these labels.
\end{revcomment}
\begin{revresponse}
    To address this comment, we have edited the figure and added labels.
\end{revresponse}

\begin{revcomment}
    What is the interpretation of a negative scTOP score? Some of the plots contain small negative values. For example, in Figure 2 when plotting Ciliated vs AT2, cells high along the Ciliated axis become increasingly negative along the AT2 axis. Are these cells more or less like AT2 cells than the zero point? A similar result is seen in Figure 7 where high AT1 axis cells become increasing negative along the AT2 axis. Some other moderately high negative values are also seen in Figure 4 and 6.
\end{revcomment}
\begin{revresponse}
    To address the interpretation of negative scores, we have added the following section to the Supplementary Information:
    \begin{changes}
        \subsection*{Interpreting scTOP scores}
        In theory, the attractor-network order parameters we have adapted for gene expression data range between negative one and positive one. An attractor state order parameter with a value of positive one means the system is perfectly aligned with that particular attractor state, and negative one means it is the exact opposite of that attractor state. Zero means that it is neither correlated nor anti-correlated with that attractor state.

        In practice, scTOP scores generally do not reach positive one. scRNA-seq data contains significant numbers of dropouts, and these zeroes decrease the magnitude of the scTOP score. When an aggregate score is taken, the effect of dropouts is decreased and the scTOP score increases. This is important to keep in mind when interpreting scTOP scores; any score above 0.1 is notable, and since the data is so noisy even the best-case score may be well below 1. It's helpful to have a control population to compare with, to have a measure of the range of best-case scTOP scores. This is also why it's best to compare distributions of scores rather than scores for individual cells, since noise plays a non-negligible role.

        scTOP scores are rarely negative. Negative values for scTOP scores correspond to cases where the sample gene expression profile is the opposite of the reference profile. For example, a score of -1 would indicate that all the genes which are highly-expressed in the reference are lowly-expressed in the sample, and vice versa. There are cases where the scores are low, negative values, indicating that the sample does not align well with the reference and that some genes which are highly- (lowly-) expressed in the reference are lowly- (highly-) expressed in the sample. From the data, it appears that it’s unlikely for gene networks to contain cell types which are opposites of one another. One possible reason is that some genes, like housekeeping genes, are similarly expressed across all cell types. Also, it’s possible that cells which expressed the exact opposite genes of any given cell type would not be able to survive.
    \end{changes}
\end{revresponse}

\begin{revcomment}
    Are there any dataset-specific preprocessing steps done? For example filtering cells or genes based on quality control metrics. If so, they 
should be reported so that the results are reproducible.
\end{revcomment}
\begin{revresponse}
    We took the datasets as they were deposited and did not filter any genes. As described in the Supplementary Information section ``Reference basis construction,'' we discarded cells if there were fewer than 100 cells of the corresponding cell type. This is because the reference basis needs at least 100 cells per cell type in order to represent accurate expression profiles for the given cell types.
\end{revresponse}

\begin{revcomment}
    The code in the linked Github repository contains a tutorial, but the
 code used to run scTOP on the other datasets should also be provided.
\end{revcomment}
\begin{revresponse}
    We have made the following changes to address this comment:

    \begin{changes}
        We have implemented scTOP in an easy-to-use Python package, available on the \href{https://pypi.org/project/scTOP/}{Python Package Index} and \href{https://github.com/Emergent-Behaviors-in-Biology/scTOP}{Github}. The code for the analyses in this paper is available in a separate \href{https://github.com/Emergent-Behaviors-in-Biology/scTOP-manuscript}{Github repository}.
    \end{changes}
\end{revresponse}

\printbibliography

\end{document}
